Providing a Unified Software Environment for Canada’s National Advanced Computing Centers

### 2019-04-10 paper.tex
Changed wiki URL
Merged Ryan's comments.

The second principle is that the software environment should be decoupled from the underlying operating system. Since each tool for software package installation has strengths and weaknesses, we adopted a layered approach to leverage the best tool for each kind of software package. We briefly describe this in Section~\ref{sec:Layered_environment}.

@@

\subsection{CVMFS architecture}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}. There is one source of truth, the stratum 0 server, where software is injected.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.

Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed. Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in automatic deduplication. Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and invalidate their local caches when content is changed, by means of a Merkle tree.

@@

We also have a production repository, hosted on the production servers. By design, software packages are always copied to the production repository from the development repository, and never directly from the build infrastructure. This step also minimizes the risk of faulty software affecting end users. This repository holds all of the software environment which is visible to the end users on all of our clusters. Once software is published to this repository, it becomes visible on all clusters within about 10 minutes.

@@

One additional use case for CVMFS is to distribute datasets. By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different  configurations and access methods are needed for data repositories. At the time of writing this paper, this is used by some of our research groups in genomics, but not directly by our staff.

@@

The second layer is what we call our compatibility layer. It contains almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.

The third layer is the scientific application layer. It contains mostly scientific or performance critical applications, including for example MPI implementations, the CUDA toolkit, mathematical libraries, and domain specific applications.

\begin{figure*}
  \begin{center}
  {\Large
  \begin{tabular}{|p{0.7\textwidth}|}
  \hline \cellcolor{cyan}
  EasyBuild layer: modules for Intel, PGI, Open MPI, MKL, high-level applications.
  Multiple instruction sets (SSE3, AVX, AVX2, AVX512).\newline
  \texttt{/cvmfs/soft.computecanada.ca/easybuild/\{modules,software\}/2017}\\
  \hline \cellcolor{green}
  EasyBuild-generated modules around Nix profiles: for example Clang, Eclipse,
  Erlang, FPC, GCC, Ruby.\newline
  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/[a-z]\**}\\
  \hline \cellcolor{yellow}
  Nix layer: GNU C Library (glibc), autotools, make, bash, cat, ls, awk, grep, etc. \newline
  \texttt{module load nixpkgs/16.09 $\Rightarrow$ \$EBROOTNIXPKGS=} \newline
  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/16.09}\\
  \hline \cellcolor{lightgray} 
  Gray area: Lustre client libraries, IB/OmniPath/InfiniPath client libraries 
  (all dependencies of Open MPI). In Nix layer, but can be overridden using
  \texttt{PATH} and \texttt{LD\_LIBRARY\_PATH}.\\
  \hline \cellcolor{pink}
  OS kernel, daemons, drivers, libcuda, anything privileged (e.g. the sudo
  command): always local. Some legally restricted software too, such as VASP.\\
  \hline
  \end{tabular}
  }
  \end{center}
  \caption{Layered structure of our software stack. The bottom layer (pink) is entirely managed by the system administrators of each cluster. The second layer (grey) contains libraries that can be installed centrally, through Nix, but can be overridden locally. The third layer (yellow) contains everything that one would typically find in the Linux user space. The fourth layer (green) contains packages that you would typically find in an operating system, but for which multiple versions are required. They are installed through Nix, but wrapped into a module by EasyBuild. The top layer (cyan) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
  \label{tab:layers}
\end{figure*}

\section{Compatibility layer}
\label{sec:Compatibility_layer}
Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would provide. We exclude the Linux kernel and its modules, hardware drivers, and privileged binaries from this layer.

@@

The installation path of each software package installed through Nix resides in a path called the Nix store. This store contains a single directory for each version of each software package installed through Nix. These directories are named after a hash calculated from the installation recipe and its dependencies. This creates a very large number of subdirectories, many of which contain almost identical files.

In addition to the store, Nix creates directories that are called profiles. These profile directories contain a forest of symbolic links that point to files in the Nix store. For example, it contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store. A profile directory also contains \texttt{lib/} and \texttt{include/} directories which contain symbolic links to libraries and header files located in the store.

Profiles are the paths that are exposed to the end users, which is done by adding paths to their \texttt{PATH}, \texttt{CPATH} and other similar environment variables. Each profile ever created is kept, but users are only exposed to the most recent one. Each profile represents the complete state of the user-exposed environment at a given time.

@@

This works well until a software package exposes paths to the Nix store directly, instead of the paths to Nix profiles. We call this a Nix store leak. This happens for example with Python when using virtual environments. Virtual environments in Python copy the \texttt{python} executable to a secondary location, outside of Nix. However, all binaries installed by Nix use \texttt{RPATH} for shared object resolutions in the store directly. This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself. Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary. We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed through EasyBuild instead.

@@

In general, Nix provides an environment that is internally consistent but is best developed using Nix itself, through tools such as \texttt{nix-shell}. Such tools can only be used if the Nix store is writable, an impossibility with CVMFS. Attempting to use Nix in conjunction with other development tools using environment variables can be fragile at times. Moreover, the dependency of the hash on the whole installation recipe, instead of only the dependencies (as, for example, implemented in Spack~\cite{Spack}), causes the store path to change even if a bug fix or security update needs to be applied, aggravating this problem. We must note however, that some clusters in the Compute Canada provide Nix with a writable Nix store, which is completely separate from the software stack we are describing here.

@@

The framework contains all of the core code which drives the logic of the entire software installation process, including parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.

@@

\section{Making the software environment available on your system}
\label{sec:Making_it_available}
Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM. This is described on this webpage:
\begin{center}
    \url{https://docs.computecanada.ca/wiki/Accessing_our_software_environment_on_your_computer}
\end{center}

After CVMFS is configured, one can simply execute this command in order to set up Compute Canada's software environment:

@@

@@

If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers. If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us as described in the web page mentioned before. Note that this software stack comes with absolutely no warranty.

>>>>

The second principle is that the software environment should be decoupled from the underlying operating system. Since each tool for software package installation has different strengths and weaknesses, we adopted a layered approach to leverage the best tool in each situation. We briefly describe this in Section~\ref{sec:Layered_environment}.

@@

\subsection{CVMFS architecture}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}. There is one source of truth, the stratum 0 server, where software is published.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that publishing new files is not possible until it is back online.

Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed. Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in automatic deduplication. Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and quickly switch their repositories to new revisions when updates are published,  by means of a Merkle tree.

@@

We also have a production repository, hosted on the production servers. By design, software packages are always copied to the production repository from the development repository, and never directly from the build infrastructure. This step also minimizes the risk of faulty software affecting end users. This repository holds the software environment which is visible to the end users on all of our clusters. Once software is published to this repository, it becomes visible on all clusters within about 10 minutes.

@@

One additional use case for CVMFS is to distribute datasets. By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different configurations and access methods are needed for data repositories. At the time of writing this paper, data repositories are used by some of our research groups in genomics, but not directly by our staff yet.

@@

The second and third layers constitute what we call our compatibility layer, containing almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.

The fourth and fifth layers are the scientific application layers. They contain mostly scientific or performance critical applications, including for example MPI implementations, the CUDA toolkit, mathematical libraries, and domain specific applications.

\begin{table*}
  \begin{center}
  {\Large
  \begin{tabular}{|p{0.7\textwidth}|}
  \hline \cellcolor{cyan}
  EasyBuild layer: modules for Intel, PGI, Open MPI, MKL, high-level applications.
  Multiple instruction sets (SSE3, AVX, AVX2, AVX512).\newline
  \texttt{/cvmfs/soft.computecanada.ca/easybuild/\{modules,software\}/2017}\\
  \hline \cellcolor{green}
  EasyBuild-generated modules around Nix profiles: for example Clang, Eclipse,
  Erlang, FPC, GCC, Ruby.\newline
  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/[a-z]\**}\\
  \hline \cellcolor{yellow}
  Nix layer: GNU C Library (glibc), autotools, make, bash, cat, ls, awk, grep, etc. \newline
  \texttt{module load nixpkgs/16.09 $\Rightarrow$ \$EBROOTNIXPKGS=} \newline
  \texttt{/cvmfs/soft.computecanada.ca/nix/var/nix/profiles/16.09}\\
  \hline \cellcolor{lightgray} 
  Gray area: Lustre client libraries, IB/OmniPath/InfiniPath client libraries 
  (all dependencies of Open MPI). In Nix layer, but can be overridden using
  \texttt{PATH} and \texttt{LD\_LIBRARY\_PATH}.\\
  \hline \cellcolor{pink}
  OS kernel, daemons, drivers, libcuda, anything privileged (e.g. the sudo
  command): always local. Some legally restricted software too, such as VASP.\\
  \hline
  \end{tabular}
  }
  \end{center}
  \caption{Layered structure of our software stack. The bottom layer (pink) is entirely managed by the system administrators of each cluster and is not on CVMFS. The second layer (grey) contains libraries that are available via Nix on CVMFS, but can be overridden locally. The third layer (yellow) contains everything that one would typically find in the Linux user space. The fourth layer (green) contains packages that you would typically find in an operating system, but for which multiple versions are required. They are installed through Nix, but wrapped into a module by EasyBuild. The top layer (cyan) contains packages that are not typically installed in a Linux distribution, and are installed via EasyBuild.}
  \label{tab:layers}
\end{table*}

\section{Compatibility layer}
\label{sec:Compatibility_layer}
Isolating the software environment from the operating system requires a compatibility layer containing almost everything that an OS would provide. We exclude the Linux kernel and its modules, hardware drivers, and privileged binaries from this layer. However, some libraries such as InfiniBand client libraries that are close to the hardware are also available in this layer (the gray area in Table~\ref{tab:layers}) and can optionally be used, or can be provided locally outside CVMFS.

@@

The installation path of each software package installed through Nix resides in a path called the Nix store. This store contains a single directory for each version of each software package installed through Nix. These directories are named after a hash calculated from the installation recipe and its dependencies. This creates a very large number of subdirectories, many of which contain nearly identical files.

In addition to the store, Nix creates directories that are called profiles. These profile directories contain a forest of symbolic links that point to files in the Nix store. For example, each profile contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store. A profile directory also contains \texttt{lib/} and \texttt{include/} directories which contain symbolic links to libraries and header files located in the store. We provide one main profile as the compatibility layer and some custom profiles for applications from that layer for which we provide multiple versions via modules that are generated by EasyBuild. Those custom profiles hence provide a hybrid between the top layer and the compatibility layer.

Profiles are the paths that are exposed to the end users, which is done by adding paths to their \texttt{PATH}, \texttt{CPATH} and other similar environment variables. Each profile ever created is kept, but users are only exposed to the most recent one. Each profile represents the complete state of the user-accessible environment at a given time.

@@

This works well until a software package exposes paths to the Nix store directly, instead of the paths to Nix profiles. We call this a Nix store leak. This happens for example with Python when using virtual environments. Virtual environments in Python copy the \texttt{python} executable to a secondary location, outside of Nix. However, all binaries installed by Nix use \texttt{RPATH} for shared object resolution directly in the store. This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself. Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary. We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed through EasyBuild instead.

@@

In general, Nix provides an environment that is internally consistent but is best developed using Nix itself, through tools such as \texttt{nix-shell}. Such tools can only be used if the Nix store is writable, an impossibility with CVMFS. Attempting to use Nix in conjunction with other development tools using environment variables can be fragile at times. Moreover, the dependency of the hash on the whole installation recipe, instead of only the dependencies (as, for example, implemented in Spack~\cite{Spack}), causes the store path to change even if a bug fix or security update needs to be applied, aggravating this problem. We must note however, that some clusters in Compute Canada provide Nix with a writable Nix store, which is completely separate from the software stack described here.

@@

The framework contains the core code which drives the logic of the entire software installation process, including parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.

@@

\section{Making the software environment available on your system}
\label{sec:Making_it_available}
To access the Compute Canada repositories on your own computer, you can install and configure the CVMFS client as described here:
\begin{center}
    \url{https://docs.computecanada.ca/wiki/Accessing_CVMFS}
\end{center}

Then, to set up the Compute Canada software environment, simply execute this command:

@@

If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, or to present our software environment to your users, we ask that you first contact us as described in the documentation web page above. Also, if you set up a large number of client nodes we require that you deploy them behind one or more caching HTTP proxies, to avoid overloading the stratum 1 servers. Note that this software stack is provided as is and comes with absolutely no warranty.

@@
